---
title: "Install"
format: html
editor: visual
minimal: true
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
vignette: >
  %\VignetteIndexEntry{Install}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

For manual installation, an [R](https://www.r-project.org/) version \>= 4.6.0 is
required.

# Running workflows in the cloud

No installation is needed to run the workflows: Metabonaut is available as a
*Curated Workshop* on the [Bioconductor Workshop
platform](https://workshop.bioconductor.org), a Galaxy-based service hosted on
the Jetstream2 academic cloud (NSF-ACCESS).

-   Log in on [workshop.bioconductor.org](https://workshop.bioconductor.org)
    with a Bioconductor account.
-   Launch the **Metabonaut** workshop. This starts a private RStudio session,
    based on the same docker image as below, with all required packages and the
    example data already installed.
-   Open any of the Quarto files in the *vignettes* folder and evaluate the R
    code blocks.

Note that these sessions are ephemeral: all data and progress are erased when
the workshop is stopped, so download a local copy of anything you want to keep.

# Running workflows locally

To install all the packages required to run the workflows locally run the
code as follow in an R session (R >= 4.6.0):

```{r eval=FALSE}
install.packages("BiocManager")
BiocManager::install(
    "RforMassSpectrometry/MsIO@a669fa1303a023161581b7a16caed3ed20a43299",
    ask = FALSE, dependencies = TRUE)

BiocManager::install("RforMassSpectrometry/Metabonaut",
                     dependencies = TRUE, ask = FALSE, update = TRUE)
```

If you get some error message when downloading form GitHub it might be due to an
expired token. Remove them using this code below and try to install again:

```{r eval=FALSE}
#See if any token is there:
gitcreds::gitcreds_get()

# Delete them:
gitcreds::gitcreds_delete()
```

The workflow source files (quarto and Rmarkdown documents) can be downloaded
from Metabonaut's GitHub
[repository](https://github.com/RforMassSpectrometry/Metabonaut) (download
either individual files from the *vignettes* folder, or the full repository).

# Docker image

The vignettes files along with an R runtime environment including all required
packages and data as well as the RStudio (Posit) editor are all bundled in a
*docker* container.

After installation, this docker container can be run on the computer and the
code and examples from the vignettes can be evaluated within this environment
(without the need to install any additional packages or files).

-   If you don't already have, install [docker](https://www.docker.com/). Find
    installation information [here](https://docs.docker.com/desktop/).
-   Get the [docker
    image](https://hub.docker.com/r/rformassspectrometry/metabonaut) of this
    tutorial e.g. from the command line with:

```
docker pull rformassspectrometry/metabonaut:RELEASE_3_23
```

::: {.callout-note appearance = "minimal"}
ℹ️ the tag *RELEASE_3_23* selects the docker image specific for Bioconductor
release 3.23. Docker images for upcoming Bioconductor releases will be made
available using a different tag (e.g. RELEASE_3_24) ensuring this version of the
image will be available also in future.
:::

-   Start the docker container, either through the Docker Desktop, or on the
    command line with

```
docker run -e PASSWORD=bioc -p 8787:8787 rformassspectrometry/metabonaut:RELEASE_3_23
```

-   Enter [`http://localhost:8787`](http://localhost:8787) in a web browser and
    log in with username `rstudio` and password `bioc`.
-   In the RStudio server version: open any of the Quarto files in the
    *vignettes* folder and evaluate the R code blocks in that document.
