---
title: "Normalization and feature selection with the notame package"
format: html
tbl-cap-location: bottom
editor: source
minimal: true
bibliography: notame-references.bib
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
vignette: >
  %\VignetteIndexEntry{Complete end-to-end LC-MS/MS Metabolomic Data analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Introduction
Herein, we will perform normalization and feature selection using the
*notame* R packages, originally developed in parallel with a protocol
article published in the "Metabolomics Data Processing and Data
Analysis—Current Best Practices” special issue of the Metabolites journal
[@klavus_notame_2020]. The main outcome is identifying interesting features
for laborious downstream steps relating to biological context, such as
annotation and pathway analysis, which fall outside the purview of *notame*.
The associated protocol article presents a sequence of steps that is easily
adopted by new practitioners. We will not follow the protocol article
exactly and instead do an analysis which meshes with the trunk of the end-to-
end Metabonaut workflow.

The normalization process is based on QC samples so the sample size here is limiting, with only four QC samples. The QC samples were also pooled from another sample collection, whereas typically, the QC samples would be pooled from the same sample collection. These limitations will become evident below, so we make an effort to explain the basic things one would want to look out for.

# Setup
Let's prepare by attaching packages and loading the preprocessed data as a
SummarizedExperiment object returned from `MSExperiment::quantify()`.

```{r, message=FALSE, warning=FALSE}
library(alabaster.se) # Portable loading and saving of SummarizedExperiment
library(BiocParallel) # Parallelization
library(cowplot) # Combinine plots in pane
library(limma) # Differential expression analysis
library(notame) # Preprocessing
library(notameViz) # Visualization
library(notameStats) # Feature selection
library(SummarizedExperiment) # The SummarizedExperiment container

# Set up parallel processing using 2 cores
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(10))
} else {
    register(SnowParam(2))
}

dir <- system.file("extdata", "preprocessed_res", package = "Metabonaut")
se <- readObject(dir)
se
```
The `SummarizedExperiment` container supports several assays. In *notame*, you
need to specify the assay using the `assay.type` parameter if using multiple
assays. Herein, we have two assays: "raw" contains the detected peaks
without gap-filling, whereas "raw_filled" contains the gap-filled peak
table. For this demo to go smoothly, we prepare the SummarizedExperiment
object for all the functions at once, mostly by renaming and creating
columns.

```{r, message=FALSE, warning=FALSE, results='hold'}
# Rename columns of sample data
rename_ind <- which(colnames(colData(se)) %in%
                    c("sample_name", "sample_type", "injection_index"))
colnames(colData(se))[rename_ind] <- c("Sample_ID", "QC", "Injection_order")
# Change rownames to be identical with "Sample_ID" column
colnames(se) <- colData(se)$Sample_ID
# Change name of pooled samples to "QC" in the "QC" column
se$QC[se$QC == "pool"] <- "QC"
# Convert phenotype column to factor
se$phenotype <- as.factor(se$phenotype)

# In feature data, create Feature_ID column from rownames
rowData(se) <- cbind(Feature_ID = rownames(se), rowData(se))
# Create "Split" column with analytical mode
rowData(se)$Split <- "HILIC_pos"
# Create "Flag" column for flagging low-quality features
rowData(se)$Flag <- NA

# Convert assay from DelayedMatrix to regular matrix
assay(se, "raw_filled") <- as.matrix(assay(se, "raw_filled"))
assay(se, "raw") <- as.matrix(assay(se, "raw"))
```

# Normalization
Features with a low detection rate are flagged, as they are deemed too
unreliable not only for statistical analysis, but also for the normalization
process which relies heavily on QC samples. Flagged features are
automatically excluded in most functions in *notame*, but can be included using
`all_features = TRUE`. We set the detection rate threshold for QC samples at
70% [@broadhurst_guidelines_2018], plus a within-group threshold of 80%. The
within-group threshold must be met in at least one study group.

```{r, message=FALSE, warning=FALSE, results = 'hold'}
se <- flag_detection(se, qc_limit = 0.7, group_limit = 0.8,
                     group = "phenotype", assay.type = "raw_filled")
```

Next, we correct for drift using a cubic spline, relating each
features' abundance in QC samples to injection order
[@kirwan_characterising_2013]. Drift correction is performed in log-space
since the log transformed data better follows assumptions of cubic spline
regression. The value for the smoothing parameter is, by default, optimized
using leave-one-out cross-validation to avoid overfitting. The abundances
are corrected by adding the mean of a feature’s abundance in the QC samples
and subtracting the predicted fit for each feature.

```{r, message=FALSE, warning=FALSE}
se <- correct_drift(se, assay.type = "raw_filled", name = "drift_norm")
```

Brief notes about drift correction are stored in the `DC_note` column of
feature data. For example, it is noted if drift correction couldn't be
performed because because at least four QC samples with values are needed
for fitting the cubic spline.

Next, we apply probabilistic quotient normalization (PQN) to reduce unwanted
variation from differential dilution of samples
[@dieterle_probabilistic_2006]. The central challenge here is to reduce
unwanted variation from dilution, whether biological or experimental, while
accounting for the possibility of biologically meaningful variation in total
feature abundances. In probabilistic quotient normalization, the most
probable dilution factor is determined for each sample as the median of
quotients calculated for each feature relative to the median of each
feature's abundance in reference samples. The sample abundances are then
divided by the dilution factor. Below, we use quality features in QC samples as
reference, although the literature suggests that the choice of reference
samples isn't critical [@dieterle_probabilistic_2006].

```{r, message=FALSE, warning=FALSE}
se <- pqn_normalization(se, ref = "qc", method = "median",
                        assay.type = "drift_norm", name = "dil_norm")
```

Next, let's visualize the data before and after normalization for drift and
dilution using the *notameViz* package. Typically, the data would be assessed
visually after each step of the normalization process, using the
`save_QC_plots()` wrapper to save relevant plots. Herein, the data is
visualized side-by-side after normalization for both drift and dilution for
conciseness. The are many parameters for customizing the visualizations, but
here we stick to the basics of coloring, shaping and ordering the samples to
best represent what is of interest.

Typically, drift in the biological samples would be evidenced by an
overabundance of low p-values in histograms relating each features'
abundance to injection order. After drift correction, the distributions
would be expected to be more uniform for all samples and biological samples.
In the case of QC samples, an overabundance of high p-values is expected
given that the QC samples are identical.

The QC samples behave as expected, but it seems that the small number of
samples does not allow us to fit linear models relating abundance to
injection order properly for all samples and biological samples, as
evidenced by the overabundance of high p-values in the top histograms.

```{r, results = 'hide'}
#| fig-cap: >
#|   Figure 1. P-values from linear regression models relating each
#|   feature to injection order. The dashed red lines represent the expected
#|   uniform distribution. A) Before normalization, featuring all samples,
#|   biological samples and QC samples. B) After normalization, featuring all
#|   samples, biological samples and QC samples
#| fig-width: 16
#| fig-height: 7
#| out-width: "100%"
#| warning: false
plot_grid(plot_injection_lm(se, assay.type = "raw_filled"),
          plot_injection_lm(se, assay.type = "dil_norm"),
          labels = "AUTO")
```

Sample boxplots are useful for inspecting the distributions of feature
abundances. One would expect the QC samples to be more alike after
normalization for drift and dilution. Depending on the instrumentation and
experimental details, drift may be apparent as a global trend or the effect of
drift is cancelled out.

In this case, the distributions of feature abundances are only slightly shifted
after normalization. The QC samples appear not to have been affected by PQN.
No obvious differences can be seen within study groups or between study
groups either. PQN generally decreased the abundances of
the biological samples to levels more similar to the QC samples. This comes
from a dilution factor of over one for each of the biological samples,
reflecting the different source and processing of QC samples.

```{r, warning=FALSE}
#| fig-cap: >
#|   Figure 2. Boxplots representing the distributions of feature
#|   abundances in each sample by injection order, featuring the median as a
#|   black line, the interquartile range as a box and the 1.5x the
#|   interquartile range as whiskers. A) Before normalization. B) After
#|   normalization.
#| fig-width: 15
#| fig-height: 7
#| out-width: "100%"
plot_grid(plot_sample_boxplots(se, order_by = "Injection_order",
                               fill_by = "phenotype",
                               assay.type = "raw_filled"),
          plot_sample_boxplots(se, order_by = "Injection_order",
                               fill_by = "phenotype", assay.type = "dil_norm"),
          labels = "AUTO")
```

For the remaining visualizations which all involve Euclidean distances, the
data was autoscaled. Conservative interpretation focusing on the QC samples is
enough to assess the normalization process, but intuitions about the data at
large can be formed based on the biological samples.

In distance density plots, one would generally expect a single prominent
density peak for QC samples and greater separation between QC samples and
biological samples after normalization, which also is the case herein. The
separation between QC samples and biological samples has not increased,
although in this case it may be an artifact from the sample size affecting the
kernel density estimate.

```{r}
#| fig-cap: >
#|   Figure 3. Density plot of Euclidean distances between samples. A) Before
#|   normalization. B) After normalization.
#| fig-width: 16
#| fig-height: 7
#| out-width: "100%"
plot_grid(plot_dist_density(se, assay.type = "raw_filled", title = NULL),
          plot_dist_density(se, assay.type = "dil_norm", title = NULL),
          labels = "AUTO")
```

PCA is useful for visualizing the similarity of individual samples and groups
as it represents the data along axes which, consecutively, maximally explains
the variation. PCA is especially useful in an exploratory sense, allowing us to
relate covariates to the grouping of samples. Typically, the QC samples will
group more tightly after normalization and the overall structure is preserved.
The variation explained by the first principal components can also be expected
to increase as normalization should reduce the masking effect of unwanted
variation.

As expected, the overall structure is intact and the QC samples group more
tightly, indicating that we have reduced unwanted variation. The CVD group
seems to group more tightly after normalization; the opposite can be
said for the CTR group. Such local differences could be investigated further
using t-SNE and coloring for injection order. There is little difference in the
variance explained by the first two components.

```{r, results=FALSE}
#| fig-cap: >
#|   Figure 4. PCA plots of samples, shape by study group and color by
#|   injection order. A) Before normalization. B) After normalization.
#| fig-width: 10.5
#| fig-height: 4
#| out-width: "100%"
plot_grid(plot_pca(se, color = "Injection_order",
                   shape = "phenotype", assay.type = "raw_filled"),
          plot_pca(se, color = "Injection_order",
                   shape = "phenotype", assay.type = "dil_norm"),
          labels = "AUTO")
```

PCA and hierarchical clustering using Ward's criterion are complementary
unsupervised methods operating in Euclidean space. While PCA focuses on
explaining the variation with maximally reduced dimensionality, hierarchical
clustering allows observation of relationships between samples at higher
resolution by minimizing within-cluster variation in the clustering of samples.
Tighter clustering of QC samples is expected. Hierarchical clustering is generally more sensitive to noise than PCA.

In the present case, the QC samples appear more dissimilar after
normalization, so we may actually have ended up increasing unwanted variation.

```{r}
#| fig-cap: >
#|   Figure 5. Dendrograms of hierarchical sample clusters using Ward’s
#|   criterion on Euclidean distances between samples. A) Before normalization.
#|   B) After normalization.
#| fig-width: 20
#| fig-height: 10
#| out-width: "100%"
plot_grid(plot_dendrogram(se, color = "phenotype", assay.type = "raw_filled",
                          title = NULL, subtitle = ""),
          plot_dendrogram(se, color = "phenotype", assay.type = "dil_norm",
                          title = NULL, subtitle = ""),
          labels = "AUTO")
```

Heatmaps give insight into the actual distances between samples, where the
distances between QC samples should be smaller after normalization. Organizing
the heatmap as per hierachical sample clusters using Ward's criterion
facilitates interpretation. The QC samples should appear as a darker, more
uniform square.

In unison with the density plots, we can observe less distance between QC
samples, although the less uniform QC sample block suggests that all QC samples
did not respond equally to the normalization process.

```{r}
#| fig-cap: >
#|   Figure 6. Heatmaps of Euclidean distances between samples, grouped
#|   by hierarchical clusters using Ward's criterion. A) Before normalization.
#|   B) After normalization.
#| fig-width: 20
#| fig-height: 10
#| out-width: "100%"
plot_grid(plot_sample_heatmap(se, group = "phenotype",
                              assay.type = "raw_filled",
                              title = NULL, subtitle = ""),
          plot_sample_heatmap(se, group = "phenotype", assay.type = "dil_norm",
                              title = NULL, subtitle = ""),
          labels = "AUTO")
```

# Feature prefiltering

We flag low-quality features, excluding them from downstream steps.
In addition to the D-ratio and detection rate in the trunk of the workflow, we
flag features by their relative standard deviation in QC samples. The non-
parametric, robust versions of the D-ratio and relative standard deviation are
used[@broadhurst_guidelines_2018].

```{r, message=FALSE, warning=FALSE}
se <- flag_quality(se, assay.type = "dil_norm",
                   condition = "RSD_r < 0.2 & D_ratio_r < 0.4")

knitr::kable(flag_report(se))
```

Finally, we impute missing values. Using random forest imputation, we can
accommodate the possibility of missing values arising not only from the limit
of detection, but also because of issues in gap-filling. We drop QC samples so
as not to bias the imputation. The strict thresholds for quality metrics and
detection rate allow us to be more confident in our imputed values.

The data is now ready for feature selection with differential abundance
analysis. It's handy to save pretreated data in a separate object and continue
with a single assay for the remainder of the analysis. We also drop flagged
features, although low-quality features can be needed in annotation when
searching for specific ions or fragments of known molecules.


```{r, message=FALSE, warning=FALSE, results='hold'}
set.seed(2024)
pretreated <- impute_rf(drop_qcs(se), assay.type = "dil_norm",
                        name = "imputed")
base <- drop_flagged(pretreated)
assays(base)[names(assays(base)) != "imputed"] <- NULL
```

The out-of-bag error is promising. Now the data is ready for feature selection.

# Differential abundance analysis
Let's get a final look at the data with PCA, this time without QC samples and
coloring by the only available clinical covariate, age.

```{r, results=FALSE}
#| fig-cap: >
#|   Figure 7. PCA plot of biological samples, shape by study group and color
#|   by age.
#| fig-width: 7
#| fig-height: 7
#| out-width: "100%"
plot_pca(base, color = "age", shape = "phenotype")
```

We'll use linear models to find interesting features, adjust for false
positives from multiple testing using the false discovery rate approach and
plot the results in histograms to assess validity of the tests and get a feel
for the results. The formula interface is used in most univariate statistics
functions in *notameStats* for flexibility. By default, linear models are fit
for each feature but correction for multiple testing is only applied for
quality features. Here we also use `join_rowData()` to add the results to the
object.

```{r, message=FALSE, warning=FALSE, results = FALSE}
#| fig-cap: >
#|   Figure 8. Linear regression p-value histograms with abundance and age as
#|   independent variables. A) p-values. B) FDR-adjusted p-values (q-values).
#| fig-width: 15
#| fig-height: 7
#| out-width: "100%"
assay(base, "log2") <- log2(assay(base))
lm_results <- perform_lm(base, assay.type = "log2",
                         formula_char = "Feature ~ phenotype + age")
base <- join_rowData(base, lm_results[, c("Feature_ID", "phenotypeCVD.p.value",
                                          "phenotypeCVD.p.value_FDR",
                                          "phenotypeCVD.estimate")])

plot_grid(plot_p_histogram(list(lm_results$phenotypeCVD.p.value)),
          plot_p_histogram(list(lm_results$phenotypeCVD.p.value_FDR)),
          labels = "AUTO")
```

The p-value histogram for looks promising; it is a relatively uniform
distribution with an overabundance of low p-values. However, there are no
significant features after correction for multiple testing. For demonstration
purposes, we'll consider the feature "FT0845", which was identified as caffeine
in the trunk of the workflow, most interesting.

After differential abundance analysis or feature selection using supervised
learning, for example, the number of significant features or a ranking cutoff
allows for manual inspection of feature-wise plots. *notameViz* includes a
variety of feature-wise visualizations adaptable to a variety of study designs.
For feature "FT0845", the abundance of CTR samples is reduced after
normalization.

```{r, results=FALSE, message=FALSE}
#| fig-cap: >
#|   Figure 10. Beeswarm plots for the lowest p-value feature. A) Before
#|   normalization. B) After normalization.
#| fig-width: 14
#| fig-height: 7
#| out-width: "100%"
plot_grid(save_beeswarm_plots(pretreated["FT0845", ], x = "phenotype",
                              color = "phenotype", save = FALSE,
                              assay.type = "raw_filled")[[1]],
          save_beeswarm_plots(base["FT0845", ], x = "phenotype",
                              color = "phenotype", save = FALSE,
                              assay.type = "imputed")[[1]],
          labels = "AUTO")
```

The results are also visualized with a variety of comprehensive visualizations.
The volcano plot below shows that feature "FT0845" is distinct in having a low
p-value and the second-largest fold-change. Manhattan plots and cloud plots
could also be used to inspect how interesting features relate to m/z and
retention time. We often co-visualize results from differential abundance
analysis with a ranking of features from supervised learning for a combined
perspective.

```{r, message=FALSE, warning=FALSE}
#| fig-cap: >
#|   Figure 11. Volcano plot of p-values from limma (negative log10 scale)
#|   related to fold-change (log2 scale).
#| out-width: "100%"
volcano_plot(base, x = "phenotypeCVD.estimate",
             p = "phenotypeCVD.p.value") +
  geom_text(data = rowData(base)["FT0845", ],
            aes(label = "FT0845"),
            vjust = -2)
```

# Conclusion
Caffeine is a mild diuretic and can concentrate blood in high doses, so a
difference in biological dilution between the study groups can be expected. The
overall structure of the data seems to be intact after normalization. On
the other hand, the QC samples were from a different source so the most
probable dilution factors may not have been determined accurately for the
biological samples and drift correction likely suffered as well.

# Session information
```{r}
sessionInfo()
```

# References
